---
title: "Homework 2: Image Compression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
library(readr)
```

## PCA Compression Function

```{r}
pca_compress <- function(x, k) {
  V <- t(x) %*% x
  eig <- eigen(V)
  U <- eig$vectors
  Z <- x %*% U
  Uk <- U[, 1:k, drop = FALSE]
  Zk <- Z[, 1:k, drop = FALSE]
  X_k <- Zk %*% t(Uk)
  err <- norm(x - X_k, type = "F")
  return(list(approx = X_k, error = err))
}

load_image <- function(path) {
  x <- read_csv(path, show_col_types = FALSE)
  data.matrix(x)
}

show_image <- function(mat, main = "") {
  image(t(apply(mat, 2, rev)), col = gray.colors(256), axes = FALSE, main = main)
}

error_curve <- function(x) {
  V <- t(x) %*% x
  eig <- eigen(V)
  U <- eig$vectors
  Z <- x %*% U
  p <- ncol(x)
  errors <- numeric(p)
  for (k in 1:p) {
    X_k <- Z[, 1:k, drop = FALSE] %*% t(U[, 1:k, drop = FALSE])
    errors[k] <- norm(x - X_k, type = "F")
  }
  list(errors = errors, U = U, Z = Z)
}
```

## Image 1

```{r fig.width=10, fig.height=7}
img1 <- load_image("image1.csv")
par(mfrow = c(3, 4), mar = c(1,1,2,1))
show_image(img1, main = "Original")
for (k in 1:10) {
  res <- pca_compress(img1, k)
  show_image(res$approx, main = paste0("k = ", k))
}
```

```{r fig.width=7, fig.height=4}
par(mfrow = c(1, 1))
ec1 <- error_curve(img1)
plot(seq_along(ec1$errors), ec1$errors, type = "b", pch = 16,
     xlab = "k (rank)", ylab = "Frobenius norm error",
     main = "Image 1: Approximation Error vs k")
```

```{r fig.width=7, fig.height=4}
plot(ec1$U[, 1], type = "l", xlab = "index", ylab = "value",
     main = "Image 1: First eigenvector (u1)")
plot(ec1$Z[, 1], type = "l", xlab = "row index", ylab = "value",
     main = "Image 1: First PC score (z1)")
```

The first eigenvector shows a sharp spike at one specific column,
corresponding to the dark vertical bar in the image, while the rest
of the eigenvector is near zero. The first PC score fluctuates
noisily across all rows, since the background texture in this image
is essentially random noise rather than smooth structure. Because
the error decreases nearly linearly with k, this image is a poor
candidate for low-rank compression — most of its variance is spread
across many components rather than concentrated in a few.

**My opinion**: Given the near-linear error curve, no small k gives
a visually adequate approximation. Even k = 10 looks almost
identical to k = 1.

## Image 2

```{r fig.width=10, fig.height=7}
img2 <- load_image("image2.csv")
par(mfrow = c(3, 4), mar = c(1,1,2,1))
show_image(img2, main = "Original")
for (k in 1:10) {
  res <- pca_compress(img2, k)
  show_image(res$approx, main = paste0("k = ", k))
}
```

```{r fig.width=7, fig.height=4}
par(mfrow = c(1, 1))
ec2 <- error_curve(img2)
plot(seq_along(ec2$errors), ec2$errors, type = "b", pch = 16,
     xlab = "k (rank)", ylab = "Frobenius norm error",
     main = "Image 2: Approximation Error vs k")
```

```{r fig.width=7, fig.height=4}
plot(ec2$U[, 1], type = "l", xlab = "index", ylab = "value",
     main = "Image 2: First eigenvector (u1)")
plot(ec2$Z[, 1], type = "l", xlab = "row index", ylab = "value",
     main = "Image 2: First PC score (z1)")
```

The first eigenvector is a smooth, monotonically decreasing ramp
across the columns, capturing the dominant large-scale gradient in
the image. The first PC score varies irregularly from row to row,
reflecting how strongly each row expresses that gradient pattern.
Notably, the approximation error drops sharply after k = 1,
indicating this image is nearly rank-1 — a single component is
enough to reconstruct it almost perfectly.

**My opinion**: k = 1 or k = 2 is already sufficient — the
approximation is visually indistinguishable from the original at
these values.

## Image 3

```{r fig.width=10, fig.height=7}
img3 <- load_image("image3.csv")
par(mfrow = c(3, 4), mar = c(1,1,2,1))
show_image(img3, main = "Original")
for (k in 1:10) {
  res <- pca_compress(img3, k)
  show_image(res$approx, main = paste0("k = ", k))
}
```

```{r fig.width=7, fig.height=4}
par(mfrow = c(1, 1))
ec3 <- error_curve(img3)
plot(seq_along(ec3$errors), ec3$errors, type = "b", pch = 16,
     xlab = "k (rank)", ylab = "Frobenius norm error",
     main = "Image 3: Approximation Error vs k")
```

```{r fig.width=7, fig.height=4}
plot(ec3$U[, 1], type = "l", xlab = "index", ylab = "value",
     main = "Image 3: First eigenvector (u1)")
plot(ec3$Z[, 1], type = "l", xlab = "row index", ylab = "value",
     main = "Image 3: First PC score (z1)")
```

The first eigenvector shows a smooth, wave-like pattern that mirrors
the overall brightness structure of the image from top to bottom.
The first PC score forms a bell-shaped curve, rising as the rows
move from the bright bedding into the darker body of the puppy and
back to lighter background. This reflects the image's large-scale,
spatially smooth intensity variation, which PCA can compress
efficiently.

**My opinion**: k ≈ 8-10 gives a visually adequate approximation —
the puppy's shape, ear position, and shading become clearly
recognizable by around k = 8, and results beyond that show
diminishing returns.

## Image 4

```{r fig.width=10, fig.height=7}
img4 <- load_image("image4.csv")
par(mfrow = c(3, 4), mar = c(1,1,2,1))
show_image(img4, main = "Original")
for (k in 1:10) {
  res <- pca_compress(img4, k)
  show_image(res$approx, main = paste0("k = ", k))
}
```

```{r fig.width=7, fig.height=4}
par(mfrow = c(1, 1))
ec4 <- error_curve(img4)
plot(seq_along(ec4$errors), ec4$errors, type = "b", pch = 16,
     xlab = "k (rank)", ylab = "Frobenius norm error",
     main = "Image 4: Approximation Error vs k")
```

```{r fig.width=7, fig.height=4}
plot(ec4$U[, 1], type = "l", xlab = "index", ylab = "value",
     main = "Image 4: First eigenvector (u1)")
plot(ec4$Z[, 1], type = "l", xlab = "row index", ylab = "value",
     main = "Image 4: First PC score (z1)")
```

The first eigenvector displays several smooth peaks and valleys
corresponding to alternating light and dark regions (e.g., rooftops
versus roads or shadows). The first PC score follows a similar
pattern with a sharp dip in the middle, likely marking a distinct
dark structure such as a shadow or building edge. The relatively
smooth structure allows the image to be compressed effectively
within a moderate number of components.

**My opinion**: k ≈ 6-7 is enough to make the main structures
(buildings, roads, shadows) clearly identifiable, though a few fine
details keep sharpening slightly up to k = 10.